---
title: "Map AquaMaps species to BC CRS from HCAF"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(oharac) ### remotes::install_github('oharac/oharac')
library(terra)
library(tidyverse)
library(here)
source(here('common_fxns.R'))

```

# Summary

Convert HCAF maps from AquaMaps into the BC 1 km x 1 km projection.

# Methods

For each species, bind the AquaMaps HCAF cell presence values with a lookup dataframe of LOICZID to BC cell ID.  Save out a vector of just the new BC ID cell values, per species (rather than per AquaMaps species ID).

## Set up directories and helper functions

```{r}
am_dir <- '/home/shares/ohi/spp_vuln/aquamaps_2021'

get_am_spp_envelopes <- function() {
  env_file <- file.path(am_dir, 'species_envelope_summary.csv')
  
  ### create a tidy summary of environmental envelopes
  if(!file.exists(env_file)) { ### create from scratch
    env_info <- data.table::fread(file.path(am_dir, 'ver10_2019_hspen.csv')) %>%
      janitor::clean_names() %>%
      rename(am_sid = species_id)

    params <- c('depth', 'temp', 'salinity', 'prim_prod', 'ice_con', 'oxy', 'land_dist')
    prm_or <- paste0(params, collapse = '|')
    dist_vals <- c('min', 'pref_min', 'mean', 'pref_max', 'max')
    env_sum_df <- env_info %>%
      rename(depth_mean = mean_depth) %>%
      select(am_sid, contains(params)) %>%
      gather(param_full, value, -am_sid) %>%
      mutate(param = str_extract(param_full, prm_or),
             dist = str_replace(param_full, paste0('(', prm_or, ')_'), '')) %>%
      select(-param_full) %>%
      spread(dist, value) %>%
      filter(yn == 1) %>%
      select(-yn) %>%
      gather(dist, value, -am_sid, -param) %>%
      mutate(dist = factor(dist, levels = dist_vals),
             value = ifelse(param == 'depth', log10(value + 1), value),
             param = ifelse(param == 'depth', 'log10_depth', param))

    write_csv(env_sum_df, env_file)
  }
  env_df <- data.table::fread(env_file) %>%
    mutate(am_sid = tolower(am_sid))
  return(env_df)
}

get_am_spp_cells <- function(occurcells_cut = 0, prob_cut = 0) {
  
  spp_cell_file <- file.path(am_dir, 'hcaf_species_native_clean.csv')
  ### file created in _setup/species/aquamaps/2_map_aquamaps_to_moll.Rmd
  
  spp_cells <- data.table::fread(spp_cell_file)
  
  if(occurcells_cut > 0) {
    spp_valid <- get_am_spp_info() %>%
      filter(occur_cells >= occurcells_cut) %>%
      dplyr::select(am_sid, sciname)
    spp_cells <- spp_cells %>%
      filter(am_sid %in% spp_valid$am_sid)
  }
  if(prob_cut > 0) {
    spp_cells <- spp_cells %>%
      filter(prob >=  prob_cut)
  }
  return(spp_cells)
}

```


## Read and examine species cell occurrences

```{r clean am spp cells}
spp_cell_file <- file.path(am_dir, 'hcaf_species_native_clean.csv')

### create a cleaner version  of spp native file for speed and size!
if(!file.exists(spp_cell_file)) {
  csq_loiczid <- get_hcaf_info() %>%
    select(loiczid, csquare_code) %>%
    distinct()
  spp_cells <- data.table::fread(file.path(am_dir, 'hcaf_species_native.csv')) %>%
    janitor::clean_names() %>%
    oharac::dt_join(csq_loiczid, by = 'csquare_code', type = 'left') %>%
    select(am_sid = species_id, loiczid, prob = probability) %>%
    distinct() %>%
    mutate(am_sid = tolower(am_sid))

  write_csv(spp_cells, spp_cell_file)
}

```

## Create a LOICZID to BC cell ID lookup

```{r create cell id lookup}
loiczid_bc_r <- rast(here('_spatial/loiczid_bc_1km.tif')) %>%
  setNames('loiczid')
ocean_bc_r <- rast(here('_spatial/ocean_bc_1km.tif')) %>%
  setNames('ocean_a')
bathy_bc_neritic <- rast(here('_spatial/bathy_bc_neritic.tif')) %>%
  setNames('neritic')
bathy_bc_shallow <- rast(here('_spatial/bathy_bc_shallow.tif')) %>%
  setNames('shallow')
cell_id_r <- loiczid_bc_r %>% setValues(1:ncell(.)) %>%
  setNames('cell_id')

### combine SpatRaster layers and convert to dataframe; drop non-ocean cells
cell_id_df <- c(loiczid_bc_r, ocean_bc_r, 
                bathy_bc_neritic, bathy_bc_shallow, 
                cell_id_r) %>%
  as.data.frame() %>%
  filter(!is.na(ocean_a))
```

```{r load am data}

am_spp_cells <- get_am_spp_cells() %>%
  filter(loiczid %in% cell_id_df$loiczid)

map_stem <- here_anx('spp_maps_bc/am_spp_bc_%s.csv')
am_sids <- read_csv(here('_data/am_spp_bc.csv')) %>%
  select(am_sid, sciname)
am_spp_info <- get_spp_info() %>%
  filter(src == 'am') %>%
  select(sciname = species, wcol, id, map_f) %>%
  full_join(am_sids, by = 'sciname') %>%
  distinct() %>%
  mutate(id = ifelse(is.na(id), str_replace_all(sciname, ' ', '_'), id)) %>%
  mutate(map_f = ifelse(is.na(map_f), sprintf(map_stem, id), map_f))

am_spp_mask_info <- am_spp_info %>%
  left_join(get_am_spp_envelopes(), by = 'am_sid') %>%
  filter(param == 'log10_depth') %>%
  # mutate(value = 10^value) %>%
  spread(dist, value) %>%
  mutate(am_sid = tolower(am_sid),
         mask   = case_when(wcol %in% c('bp', 'pel') ~ 'none', 
                              ### don't mask pelagics/benthopelagics
                            is.na(wcol)              ~ 'none', 
                              ### don't mask unknowns
                            max <=  60 ~ 'shallow',
                            max <= 200 ~ 'neritic',
                            TRUE       ~ 'none')) %>%
  mutate(mask = factor(mask, levels = c('shallow', 'neritic', 'none'), ordered = TRUE)) %>%
  select(am_sid, sciname, mask, map_f) %>%
  distinct()
  

### For species with synonyms, check that mask values of duped am_sid match up -
# x <- janitor::get_dupes(am_spp_mask_info, sciname)
### all are 'shallow'

```

``` {r process maps}
map_vec <- am_spp_mask_info$map_f %>% unique()

# spp_done <- map_vec[file.exists(map_vec)]
map_vec <- map_vec[!file.exists(map_vec)]

map_am_hcaf_to_bc <- function(out_f) {
  # out_f <- map_vec[1]
  i <- which(map_vec == out_f)
  df <- am_spp_mask_info %>%
    filter(map_f == out_f)
  s <- df$sciname %>% unique()
  
  ### identify the species ID(s) for this species
  s_sids <- df$am_sid %>% unique()
  
  if(file.exists(out_f)) {
    message('File exists for ', s, ', ... (', i, ' of ', length(map_vec), ')... skipping!')
  } else {
    message('Processing map for ', s, '... (', i, ' of ', length(map_vec), 
            '), am_sid = ', paste(s_sids, collapse = '/'), '...')
    
    
    ### identify the depth(s); if multiple, choose the least restrictive
    s_depth <- df %>% 
      filter(mask == max(mask)) %>%
      .$mask %>% unique()
    
    ### identify the cells for am_sid(s) and join to bc cells
    s_cells <- am_spp_cells %>%
      filter(am_sid %in% s_sids)
    
    if(n_distinct(s_cells$am_sid) > 1) {
      ### choose highest prob in cell for multi-am_sid spp
      s_cells <- s_cells %>%
        group_by(loiczid) %>%
        filter(prob == max(prob)) %>%
        ungroup()
    }
    s_bc <- s_cells %>%
      oharac::dt_join(cell_id_df, by = 'loiczid', type = 'left')
    
    ### filter to proper depth category
    if(s_depth == 'shallow') {
      message('... clipping ', s, ' to shallow waters only')
      s_bc <- s_bc %>%
        filter(!is.na(shallow))
    } else if(s_depth == 'neritic') {
      message('... clipping ', s, ' to neritic waters only')
      s_bc <- s_bc %>%
        filter(!is.na(neritic))
    } ### else, 'none', so no bathymetric filter
    
    s_bc <- s_bc %>%
      select(prob, cell_id)

    write_csv(s_bc, out_f)
  }
}

tmp <- parallel::mclapply(map_vec, map_am_hcaf_to_bc, mc.cores = 12)
```

