---
title: "Process nutrient pollution"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(terra)
library(sf)
library(tidyverse)
library(here)

source(here('common_fxns.R'))
source('stressor_fxns.R') ### in same dir as this Rmd

```

# Summary

Gather nutrient-related stressor layers and process to the CRS for this analysis: Mollweide, 10 km x 10 km.  Identify gapfill needs if necessary.  

# Data

All files are located on Mazu, starting from `/home/shares/ohi/` unless otherwise noted.

## Agricultural-based nutrient pollution

Nutrient pollution from OHI 2021 plume model

* This appears to be absolute amounts (tonnes) per cell, so can be normalized to an intensity (tonnes per km^2^ e.g.) to be combined with wastewater data.
* dir: `git-annex/globalprep/prs_land-based_nutrient/v2021/output/N_plume`
    * file: `pourpoints_crop_manure_leached_volt_N_2019_joined.tif`
* See OHI processing scripts for details: https://github.com/OHI-Science/ohiprep_v2021/tree/gh-pages/globalprep/prs_land-based_nutrient/v2021

## Wastewater

From the Wastewater project

* Get dataset from KNB: https://knb.ecoinformatics.org/view/doi%3A10.5063%2FF76B09.  It has been downloaded to Mazu (see below). According to metadata:
    * > Each pixel is grams N.
    * Use this note from Mel to make sure the numbers are in the same ballpark between ag and wastewater:
        * > Our model indicates that wastewater adds 6.2Tg nitrogen into coastal waters, which is approximately 40% of total nitrogen from agriculture.
        * A quick check shows that this layer results in about 4.5Tg of N, about 33% of that from agriculture-based N, so ballpark.
* dir: spp vuln mapping directory, `/stressors/nutrient/wastewater`
    * file: `global_effluent_2015_tot_N.tif`

## Aquaculture pollution

From the Food Systems project

* see https://github.com/OHI-Science/food_systems/tree/master/aquaculture/marine/
    * > These values, in the scheme of these kinds of things, should be fairly accurate (the models are pretty simple and N flows directly into the marine system). I would check the units to make sure they are the same as other datasources (I think they are tonnes/cell here).
    * Looks like the nutrient pollution from aquaculture layers are on Aurora: 
        * `/home/shares/food-systems/Food_footprint/all_food_systems/datalayers`

# Methods

For each stressor layer, reproject to 10 km Mollweide CRS and mask to ocean cells.  Layers that are in tonnes of N per cell must be converted to t/km^2^ prior to reprojecting.  Sum across all sources (verifying units carefully) then rescale based on a quantile e.g., 99.9%.

```{r define stuff}
here_ohi <- function(f = '', ...) {
  f <- paste(f, ..., sep = '/')
  f <- stringr::str_replace_all(f, '\\/+', '/')
  sprintf('/home/shares/ohi/%s', f)
}

### Load ocean proportion in Mollweide 10 km x 10 km (analysis CRS and res)
ocean_rast <- rast(here('_spatial/ocean_bc_1km.tif'))

# ### load ocean area (in km^2) in 0.1 deg WGS84 (for converting values to 
# ### intensities prior to reprojecting)
# ocean_area_0.1deg <- rast(here('_spatial/ocean_area_wgs84_0.1deg.tif'))
# 
# ### Code to create ocean area raster at 0.1 degree:                            
# base_rast_0.01deg <- raster(ext = extent(c(-180, 180, -90, 90)),
#                            res = 0.01, crs = '+init=epsg:4326')
# ocean_mask_sf <- read_sf(here('_spatial/ne_10m_ocean/ne_10m_ocean.shp'))
# ocean_rast_0.01deg <- fasterize::fasterize(ocean_mask_sf, base_rast_0.01deg)
# ocean_rast_0.1deg <- aggregate(ocean_rast_0.01deg,
#                                progress = 'text',
#                                fact = 10, fun = sum)
# ocean_rast_0.1deg <- ocean_rast_0.1deg / 100
# ocean_area_0.1deg <- ocean_rast_0.1deg * raster::area(ocean_rast_0.1deg)
# writeRaster(ocean_area_0.1deg, here('_spatial/ocean_area_wgs84_0.1deg.tif'))
```

## Nutrient pollution from OHI 2021 plume model

This is in tonnes per cell (0.5 arc-minute cell).  Convert to tonnes per km^2, aggregate, reproject to 10 km x 10 km Mollweide.  This will still be tonnes per km^2.

```{r agriculture nutrient plume data}
N_agr_out_f <- here_anx('stressors/nutrient/nutrient_agr_N_per_km2_bc.tif')
### unlink(N_agr_out_f)

if(!file.exists(N_agr_out_f)) {
  N_agr_raw_f <- here_ohi('git-annex/globalprep/prs_land-based_nutrient',
                     'v2021/output/N_plume',
                     'pourpoints_crop_manure_leached_volt_N_2019_joined.tif')
  N_agr_raw_r <- rast(N_agr_raw_f) %>%
    crop(create_bc_sf())
  # class       : SpatRaster 
  # dimensions  : 2301, 43200, 1  (nrow, ncol, nlyr)
  # resolution  : 0.008333333, 0.008333333  (x, y)
  # extent      : -180, 180, 65, 84.175  (xmin, xmax, ymin, ymax)
  # coord. ref. : lon/lat WGS 84 (EPSG:4326) 
  # source(s)   : memory
  # name        : pourpoints_crop_manure_leached_volt_N_2019_joined 
  # min value   :                                       -0.01275834 
  # max value   :                                      212.50721741 
  
  ### mask to just ocean cells to avoid land-based zeros from influencing mean
  ocean_mask_sf <- read_sf(here('_spatial/ne_10m_ocean/ne_10m_ocean.shp'))
  ocean_mask_r <- rasterize(ocean_mask_sf, N_agr_raw_r)
  N_masked_r <- mask(N_agr_raw_r, ocean_mask_r)

  ### transform to N/km^2
  area_r <- cellSize(N_agr_raw_r) / 1e6
  N_per_km2_r <- N_masked_r / area_r
  
  ### reproject to 1 km x 1 km
  N_r_bc <- project(N_per_km2_r, ocean_rast)
    

  # sum(values(N_r_bc), na.rm = TRUE)  # 47992.35
  # sum(values(N_masked_r), na.rm = TRUE) # 47552.25
  ### pretty close!
  
  ### write out to temp file for later combo with wastewater and aquaculture
  writeRaster(N_r_bc, N_agr_out_f, overwrite = TRUE)
  
}

r <- rast(N_agr_out_f)
  
plot(log10(r), axes = FALSE, 
     main = 'N pollution from ag: log(tonnes per 1 km^2 cell)')

hist(log10(r)[r > 0], main = 'distribution of log_10(N) non-zero values')

```

## Nutrient pollution from wastewater plume

These values are in grams of N per pixel.  Convert to tonnes (1e6 g = 1 t).
```{r wastewater plume data}
N_ww_out_f <- here_anx('stressors/nutrient/nutrient_wastewater_N_per_km2_bc.tif')
### unlink(N_ww_out_f)

if(!file.exists(N_ww_out_f)) {
  N_ww_raw_f <- here_anx('../spp_vuln_mapping/stressors/nutrient/wastewater',
                     'global_effluent_2015_tot_N.tif')
  N_ww_raw_r <- rast(N_ww_raw_f) %>%
    crop(create_bc_sf())
  # class       : SpatRaster 
  # dimensions  : 2406, 43200, 1  (nrow, ncol, nlyr)
  # resolution  : 0.008333333, 0.008333333  (x, y)
  # extent      : -180, 180, 65, 85.05  (xmin, xmax, ymin, ymax)
  # coord. ref. : lon/lat WGS 84 (EPSG:4326) 
  # source(s)   : memory
  # name        : global_effluent_2015_tot_N 
  # min value   :                   -617.752 
  # max value   :              170788096.000 
  
  ### Convert from grams N into tonnes N (i.e., Mg):
  N_ww_tonnes_r <- N_ww_raw_r / 1e6

  ### mask to just ocean cells to avoid land-based zeros from influencing mean
  ocean_mask_sf <- read_sf(here('_spatial/ne_10m_ocean/ne_10m_ocean.shp'))
  ocean_mask_r <- rasterize(ocean_mask_sf, N_ww_tonnes_r)
  masked_r <- mask(N_ww_tonnes_r, ocean_mask_r)
  
  ### transform to N/km^2
  area_r <- cellSize(N_ww_tonnes_r) / 1e6
  N_per_km2_r <- masked_r / area_r
  
  ### then reproject
  N_ww_bc_r <- project(N_per_km2_r, ocean_rast)
    
  # sum(values(N_ww_bc_r), na.rm = TRUE) # 27724.98
  # sum(values(masked_r), na.rm = TRUE)     # 27434.79
  ### pretty close!
  
  ### write out to temp file for later combo with wastewater and aquaculture
  writeRaster(N_ww_bc_r, N_ww_out_f)
  
}

r <- rast(N_ww_out_f)
  
plot(log10(r), axes = FALSE, 
     main = 'N pollution from wastewater: log(tonnes per 1 km^2 cell)')

hist(log10(r)[r > 0], main = 'distribution of log_10(N) non-zero values')

```

## Nutrient pollution from aquaculture

These values are in tonnes of N per pixel.

```{r aquaculture nitrogen data}
N_aq_out_f <- here_anx('stressors/nutrient/nutrient_aquaculture_N_per_km2_bc.tif')
### unlink(N_aq_out_f)

if(!file.exists(N_aq_out_f)) {
  N_aq_raw_fs <- list.files(here_anx('../spp_vuln_mapping/stressors/nutrient/aquaculture'),
                            full.names = TRUE)
  N_aq_raw_stack <- rast(N_aq_raw_fs) %>%
    crop(create_bc_sf())
  # class       : SpatRaster 
  # dimensions  : 300, 4320, 6  (nrow, ncol, nlyr)
  # resolution  : 0.08333333, 0.08333333  (x, y)
  # extent      : -180, 180, 65, 90  (xmin, xmax, ymin, ymax)
  # coord. ref. : lon/lat WGS 84 (EPSG:4326) 
  # source(s)   : memory
  # names       : marin~ientN, marin~ientN, marin~ientN, marin~ientN, marin~ientN, marin~ientN 
  # min values  :           0,           0,    0.000000,      0.0000,           0,           0 
  # max values  :           0,           0,    7.709582,    263.7675,           0,           0 

  ### mask to just ocean cells to avoid land-based zeros from influencing mean
  ocean_mask_sf <- read_sf(here('_spatial/ne_10m_ocean/ne_10m_ocean.shp'))
  ocean_mask_r <- rasterize(ocean_mask_sf, N_aq_raw_stack[[1]])
  masked_stack <- mask(N_aq_raw_stack, ocean_mask_r)
  
  ### transform to N/km^2
  area_r <- cellSize(N_aq_raw_stack[[1]]) / 1e6
  N_per_km2_stack <- masked_stack / area_r
  
  ### reproject
  N_per_km2_bc_stack <- project(N_per_km2_stack, ocean_rast)
    
  ### N per km^2 times ocean area --> total N per cell
  N_per_km2_bc_r <- sum(N_per_km2_bc_stack, na.rm = TRUE) %>%
    mask(ocean_rast)
  
  ### check by layer:
  # sum(values(masked_stack[[1]]), na.rm = TRUE)          # 0
  # sum(values(N_per_km2_bc_stack[[1]]), na.rm = TRUE) # 0
  # sum(values(masked_stack[[2]]), na.rm = TRUE)          # 0
  # sum(values(N_per_km2_bc_stack[[2]]), na.rm = TRUE) # 0
  # sum(values(masked_stack[[3]]), na.rm = TRUE)          # 2.569861
  # sum(values(N_per_km2_bc_stack[[3]]), na.rm = TRUE) # 2.644865
  # sum(values(masked_stack[[4]]), na.rm = TRUE)          # 15499.07
  # sum(values(N_per_km2_bc_stack[[4]]), na.rm = TRUE) # 15689.93
  # sum(values(masked_stack[[5]]), na.rm = TRUE)          # 0
  # sum(values(N_per_km2_bc_stack[[5]]), na.rm = TRUE) # 0
  # sum(values(masked_stack[[6]]), na.rm = TRUE)          # 0
  # sum(values(N_per_km2_bc_stack[[6]]), na.rm = TRUE) # 0
  ### full stack:
  # sum(values(masked_stack), na.rm = TRUE)       # 15501.64
  # sum(values(N_per_km2_bc_r), na.rm = TRUE)  # 11681.5
  ### not as close as I'd like - some coastal values being masked out!
  
  ### write out to temp file for later combo with wastewater and aquaculture
  writeRaster(N_per_km2_bc_r, N_aq_out_f)
  
}

r <- rast(N_aq_out_f)
  
plot(log10(r), axes = FALSE, 
     main = 'N pollution from aquaculture: log(tonnes per 1 km^2 cell)')

hist(log10(r)[r > 0], main = 'distribution of log_10(N) non-zero values')

```

## Combine N pollution and rescale

Combine nitrogen (tonnes per 1 km x 1 km cell) across agriculture, wastewater, and aquaculture sources; then rescale based on a quantile to reduce the effects of extreme outliers.  This is currently NOT log-transformed.

```{r combine all layers}
N_layers <- list.files(here_anx('stressors/nutrient'),
                       pattern = 'N_per_km2_bc.tif',
                       full.names = TRUE)
N_stack <- rast(N_layers)
N_rast <- sum(N_stack, na.rm = TRUE)

### fill southern NAs from ag rast with 0, then re-mask
values(N_rast)[is.na(values(N_rast))] <- 0
values(N_rast)[values(N_rast) < 0] <- 0 ### drop neg values
N_rast <- mask(N_rast, ocean_rast)

plot(log10(N_rast), axes = FALSE, 
     main = 'N pollution total: log(tonnes per 1 km^2 cell)')

### Rescale
# quantile(values(N_rast), c(0.9, 0.95, 0.99, 0.999, 1.0), na.rm = TRUE) %>%
#   knitr::kable()
#          90%          95%          99%        99.9%         100% 
# 0.000000e+00 1.826932e-09 8.550333e+00 4.923464e+02 4.948126e+05 

N_rast_rescale <- log_plus(N_rast) %>%
  rescale_stressor(qtile = .999)
plot(N_rast_rescale, axes = FALSE, 
     main = 'N pollution rescaled')

x <- values(N_rast_rescale)[values(N_rast_rescale) > 0]
hist(x, main = 'distribution of rescaled non-zero values (raw)')

writeRaster(N_rast_rescale, here('_data/stressors_bc/nutrient_2020.tif'), 
            overwrite = TRUE)

```

